Unilateral Divorce and the Rise of Informal Cohabitation

### 2020-09-29 draft_sub_con.tex
Edited the bibliography so it compiles on mine + temp remove "achicago" style
Edits to Sections 1 and 2

\begin{abstract}
	This paper analyzes the role of unilateral divorce for the rise of cohabitation. Exploiting the staggered introduction of unilateral divorce across the US states, we show that after the reform singles become more likely to cohabit than to marry, and that newly formed cohabitations last longer. We then provide a theoretical rationale for these facts, building a life-cycle model with partnership choice, female labor force participation and saving decisions. A structural estimation of the model suggests that unilateral divorce decreases couples' stability, which makes cohabitation preferred to couples that would have been at high risk of divorce. Since cohabiting couples formed after the reform are better matched, the average length of cohabitations increases. A counterfactual experiment reveals that the time spent cohabiting would have been halved if divorce laws had never changed.
\end{abstract}

@@

Informal cohabitation is on the rise: the share of women that ever cohabited in the United States moved from 33\% in 1987 to 60\% in 2010 \citep{manning2013}. This increase contributed to the overall changes in the structure and behavior of the American family. For example, \cite{bumpass2000} claim that the higher instability of cohabitation is responsible for the rise in the number of single mothers, which is associated with poor outcomes of children \citep{chetty2018,mclanahan2013}. Moreover, compared to married couples, the relationship-specific investments of cohabitants are lower \citep{poortman2012} and their children's well-being is worse \citep{brown2004}, even after controlling for parental resources. However, it is not clear to which extent the outcomes of cohabiting families are due to selection versus the direct effect of partnership's rights on couple's behavior. Quantifying the relative importance of these two mechanisms can be done only by understanding the rational for partnership's choices. Why do people cohabit instead of marrying? Why did cohabitation become more common over time?

This paper addresses these questions by asking what role does unilateral divorce have for the rise of cohabitation. Since marriage and cohabitation can be viewed as contracts whose attractiveness depend on their termination rights and costs, the switch from mutual consent to unilateral divorce represents a unique opportunity to learn about partnership choices. This paper answers to this question with four contributions. First, we offer empirical evidence that unilateral divorce made cohabitation more likely to be chosen and that newly formed cohabitations last longer. Second, we propose a theory of partnership choice and endogenous breakup/divorce. In particular, in our model the policy makes cohabitation the preferred choice to couples that would have experienced the highest risk of a costly divorce. Therefore, the average match quality of cohabiting couples increases, which cause cohabitation spells to last longer. Third we structurally estimate our model to reflect the empirical findings. Fourth, we perform a counterfactual experiment to assess the overall role of unilateral divorce on the rise of cohabitation and we quantify the welfare gains and losses of the reform for men and women. This paper thus consists of four parts, one for each contribution, which we now describe in more detail.

@@

In the first part of the paper we document the effect of unilateral divorce on the choice between marriage and cohabitation and on the duration of newly formed cohabitations. We use data from the National Survey of Family and the Household (NSFH) and from the National Survey of Family Growth (NSFG) to study the choice between marrying and cohabiting. Then, exploiting the exogenous variation coming from the staggered introduction of unilateral divorce over time across the US states, we estimate that newly formed couples are 5-7\% more likely to choose cohabitation over marriage compared to the pre-reform period. Interestingly, the effect's size depends on how property is divided upon divorce, being strongest in states where each spouse gets half of the wealth and where the assets shares are decided by the judge. This suggest that divorce allocations affect partnership choices.
%This suggests that the result cannot be entirely driven by divorce being a less binding commitment technology,\footnote{This mechanism is highlighted by \cite{lafortune2020}, where the equal division of assets within marriage makes it possible to efficiently invest in children even after the rise of unilateral divorce.} since the egalitarian property division rule can act as a substitute for the mutual consent regime.
Moreover, we analyze how unilateral divorce affected the duration of cohabitation spells: our estimates from a multinomial probit show that cohabitations formed after the reform last longer both because of a reduced risk of marriage and of breakup.\footnote{Henceforth we will refer to the separation from cohabitation as breakup to avoid confusion with legal separation.}

In the second part of the paper we propose a theory to understand the mechanisms underlying the facts documented in the empirical part. We build a dynamic model of intra-household decision making and search in the mating market, where agents makes decisions according to the realization of idiosyncratic income shocks, their amount of wealth and a couple-specific match quality. Single agents meet, with some probability, a potential partner with a match quality, productivity and wealth. After the draw, they decide whether to marry, cohabit or stay single. Couples make decisions about consumption, savings and female labor force participation. Women receive a productivity penalty for not working and women's time can be used to produce a public good which captures utility gains from children, durable goods and services. Cohabitation and marriage differ in their splitting costs as well as in the way property is divided when the couple dissolves. Moreover, cohabitation is subjected to stigma, which is modeled as an exogenous utility flow.\footnote{The stigma towards cohabitation is to be interpreted considering the shame towards out-out-wedlock births, premarital sex and premarital cohabitation at time of the divorce revolution.} In the case of breakup, assets are split according to individual property rights, while in the case of divorce they are divided in half.\footnote{We made this assumption because the model will be estimated using community property states data.} Breakup (unilateral divorce) can be initiated unilaterally, as opposed to mutual consent divorce, which requires the consensus of both partners. Following \cite{voena2015}, under mutual consent the couple always cooperates while married, which implies that the allocation of resources corresponds to the Pareto-efficient inter-temporal allocation. Instead, unilateral divorce (breakup) results in a lack of commitment within marriage (cohabitation),\footnote{We model the decision making in the couple building on the literature on limited commitment (see for example \cite{kk1996},  \cite{ligon2002}, \cite{marcet2019} and \cite{pavoni2018}), which has been applied to dynamic collective models in the household by \cite{mazzocco2007}, \cite{mazzocco2013}, \cite{bayot2015}, \cite{rigas2015}, \cite{voena2015}, \cite{abraham2018}, \cite{lise2018}, \cite{low2018}, \cite{foerster2019} and \cite{reynoso2019} among others.} in fact the intra-household allocation can change in response to shocks, as spouses (cohabitants) can credibly exercise the threat of divorce (breakup). The fact that couples cannot commit ex-ante to any possible division of marital surplus contributes to the creation of an imperfectly transferable utility environment, under which the Becker-Coase theorem does not hold.\footnote{For an application of an imperfectly transferable utility environment to matching and collective models see \cite{galichon2019} and \cite{weber2018}.} Hence, abandoning the mutual consent regime affects the risk of divorce and in turn the surplus of marriage.\footnote{According to \cite{becker1977}, divorce laws should not affect separation decisions \textit{``if all compensations between spouses were feasible and costless"}. The assumptions underlying the Becker-Coase theorem are discussed by \cite{chiappori2015} and \cite{fella2004}.} More precisely, the barriers to divorce---represented by its costs and rights---affect the gains of marriage with respect to cohabitation through three main channels. First, by acting as commitment technologies, they enforce a better risk sharing and a more efficient specialization in the production of public goods within marriage.\footnote{We assume that public goods are produced with money and women's time. When women stop working for producing such a good their potential productivity in the labor market decreases.} Second, they increase the risk of being ``trapped" in a bad marriage. Third, they affect the expected value of marriage by modifying the risk and value of divorce. The effects of tightening or relaxing the barriers to divorce depends on which channel prevails: for example, the introduction of unilateral divorce has an uncertain impact on the share of couples that cohabit and marry. 
Interestingly, in our framework  the outcomes of cohabitation and marriage not only depend on the rules underlying these contracts, but also on sorting on match quality. In fact, a cheap breakup is most attractive to couples whose match quality is low and couples' ability to cooperate depends on the match quality itself through its effect on the couples' stability.\footnote{It's easier to make partnership-specific investments when the risk of splitting is low, which is the case when the match quality is high.}

%
%In our framework marriage gains depend on the quality of the match: for couples whose value is low cohabitation is the best choice because breaking up is cheaper and their risk of splitting is high. Interestingly, the ability of the couple to cooperate is directly affected by the quality of the match. For this reason, the outcomes of cohabitation and marriage depend not only of the rules underlying these contracts, but the ability of the couple to cooperate is directly affected by the quality of the match.

@@

In the third part of the paper we structurally estimate the model to understand the quantitative relevance of the mechanisms that drive partnership choice. The model is estimated by indirect inference using as targets the regression results from our empirical analysis, mating market moments (NSFH), and female labor supply moments (PSID). The introduction of unilateral divorce is modeled as an unexpected policy. The estimated model closely replicates the targeted moments, and over-identification checks support the estimation results by highlighting that the model can match a number of non-targeted moments, among which the impact of unilateral divorce on cohabitation duration. According to the estimated model, a switch from mutual consent to unilateral divorce causes couples to start cohabiting more by reducing married couple's ability to cooperate and by increasing the likelihood of a costly divorce.\footnote{An increased likelihood of divorce can reduce the ability of the couple to cooperate by itself. Yet it also has a direct effect on the marriage surplus by reducing the likelihood of losing assets upon divorce. For example, if there was no wage uncertainty and women were always participating in the labor market, unilateral divorce would affect the gains of marriage via the direct effect only.} Since cohabiting couples that would have married under the older regime are better matched than the average cohabiting couple, the reform increases the stability and the duration on newly formed cohabitations. To further deepen our understanding of the mechanisms, we analyze the changes in intra-household bargaining that followed the policy change in the model. We find that the average Pareto weight of cohabiting women increases because men, fearing to lose most of their assets because of divorce, convince women to cohabit instead of marrying in exchange for more power in the couple. This mechanism is specific to community property regime: under a title-based regime men would not need to choose cohabitation to insure their assets. Consistently with empirical evidence, the impact of unilateral divorce on the likelihood to cohabit is lower in the model under a title-based regime.

The fourth and last part of the paper conducts a series of counterfactual experiments to understand the quantitative importance of the forces that contributed to the rise of cohabitation. To assess the role of unilateral divorce we perform a counterfactual experiment where unilateral divorce was never introduced. We find that people on average would have spent 1.35 years cohabiting instead of 2.26, while only 26.7\% of people would have ever cohabited instead of 34.8\%. In a second series of counterfactual experiments we find that a 20\% increase in women wages and a 20\% drop in market prices of goods increase respectively by 1\% and 2\% the share of people that ever cohabited at age 39. Both effects are driven by a reduced scope for household specialization, which is better exploited within marriage. We also perform a welfare analysis of the introduction of unilateral divorce: the possibility to cohabit limits the welfare losses for men who can secure their assets, while women suffer more because couple-specific investments like children deteriorate their value of divorcing more that for men.

\textbf{Literature.} This paper adds to three strands of the literature. First, by documenting how divorce laws influence the choice between marriage and cohabitation, we add to the existing literature that studies the effects of unilateral divorce. This policy change has been shown to affect the rate of divorce (\citealp{friedberg1998}, \citealp{wolfers2006}), female labor supply (\citealp{stevenson2008}, \citealp{voena2015}), savings (\citealp{voena2015}), marriage rates (\citealp{rasul2003,rasul2006}), children's well-being (\citealp{gruber2004}), family violence (\citealp{stevenson2006}), marriage-specific capital (\citealp{stevenson2007}), assortative mating (\citealp{reynoso2019}) and prostitution (\citealp{ciacci2017}), among the others. We complement the findings of \cite{rasul2003,rasul2006} by showing that the decrease in marriage rates after the introduction of unilateral divorce is not only driven by more people staying singles, but also by more people that choose to cohabit. This suggest that marriage and cohabitation are substitutes.\footnote{Cohabitation can also be a substitute for being single or dating, as \cite{rindfuss1990} point out. Moreover, \cite{blasutto2020} and \cite{brien2006} claim that cohabitation can also be a complement for marriage, which allows the couple to learn about its match quality before making the binding decision of getting married.} Our paper builds on \cite{voena2015}, that studies how the interaction of unilateral divorce with property rights upon divorce affected household behavior of married couples. We extend her work both by considering cohabitation as an alternative relationship and by analyzing selection into partnership. This paper also extends the work of \cite{fernandez2017} by showing that not considering cohabitation as an alternative to marriage biases upwards the negative impact of unilateral divorce on men's welfare. The intuition is that men can limit the losses stemming from the increased risk of divorce by choosing to cohabit.

Second, our paper adds to the literature that studies the choice between marriage and cohabitation. A first subset thereof has focused on identifying the gains of marriage and cohabitation, highlighting the role of commitment \citep{matouschek2008}, specialization within the couple \citep{gemici2014}, learning about match quality \citep{brien2006}, income dynamics \citep{blasutto2020}, assets \citep{lafortune2017, lafortune2020} and investment in children \citep{lundberg2015}. We extend these works by showing how an increase in the easiness of divorce not only decreases the ability of the couple to cooperate, but it also makes divorce allocations more relevant for partnership choices, since the likelihood of divorce increases. Consequently, the relative power of potential partners and the rules about division of assets upon divorce become crucial. These results highlight a new role of partners' relative power and assets for partnership choices, which is analyzed within a framework that extends the theory of \cite{blasutto2020} and \cite{gemici2014} by including saving decisions.\footnote{\cite{lafortune2020} also highlight the role of assets: our model features their intuition that assets can act as act as a commitment technology, but our framework also allows assets to influence partnership choices via a direct effect of divorce's risk. Thanks to this mechanism we can explain why unilateral divorce caused cohabitation to increase more in community property states than in title-based ones.} A second set of papers studies the effect of changes in cohabitants' rights on partnership choices and cohabiting couples' behavior, highlighting the role of alimony rights (\citealp{chiappori2017,gousse2018}), taxation \cite{leturcq2012} and division of assets at breakup (\citealp{fisher2012,gousse2018,chigavazira2019}). We extend this literature by showing that the introduction of unilateral divorce impacts both the choice to cohabit and cohabitation's stability, even though cohabitant's rights are not directly affected.\footnote{\cite{matouschek2008} study the effect of unilateral divorce on marriage rates using a framework where the choice between marriage and cohabitation is considered. Their framework does not predict changes to the propensity to cohabit.} Further, the effects on the choice to cohabit depends on property division rights, which indicates that partnership choices depend on divorce allocations. This evidence suggests that changes in family law should be designed considering that marriage and cohabitation are substitutes. 
%\cite{chiappori2017},\cite{persson2020}, \cite{leturcq2012}, \cite{gousse2018}
%Second, our paper expands the literature on cohabitation, showing how the risk of divorce is essential for understanding partnership choices. Different papers in the literature highlighted various gains of marriage with respect cohabitation: commitment \citep{matouschek2008}, labor specialization within the couple \citep{gemici2014}, learning about match quality \citep{brien2006}, income dynamics \citep{blasutto2020} and investment in children \citep{lafortune2020}. A paper closely related to ours is \cite{lafortune2020}, who highlight the role of assets as a collateral that enforces commitment and hence allows for optimal investment in children. While we think that their mechanism is important, it cannot be the whole story since we observe that the switch towards cohabitation is the strongest in states where assets are divided more equally, where their model would predict that the effect is the weakest. In this paper, we quantify how much of the shift is due to the erosion of marriage's collateral versus the role of the increased risk of divorce. We extend the work of \cite{gemici2014} and \cite{blasutto2020}, who develop collective dynamic models with limited commitment, introducing asset accumulation decisions, which are shown to play an important role for commitment and hence for partnership choice, especially when interacted with property rights upon divorce. Other papers highlighted how

Finally, this paper is tied to the large literature that studies the changes in the character of the American household over the last decades. Various studies explored the role of health improvements, wage distribution and dynamics, norms and technology on the rise in female labor force participation \citep{fernandez2004,greenwood2005,albanesi2016,greenwood2016}, the changes in household formation and dissolution \citep{greenwood2016,ciscato2019}, the rise in positive assortative mating \citep{fernandez2005,greenwood2016,ciscato2019} and the increase in the age at marriage \citep{santos2016}. We extend this literature by showing that the introduction of unilateral divorce implied an increase in cohabitation and that advances in the home production technology and the reduction in the gender wage gap also contributed to the rise in cohabitation.

The paper is organized as follows. Section 2 offers an overview of US divorce laws. Section 3 documents the effect of introducing unilateral divorce on partnership choices. Section 4 presents and develop the theoretical model. Section 5 describes the estimation of the model, while section 6 discusses the main mechanisms of the model. Section 7 reports the results of the welfare analysis. Section 8 performs a series of counterfactual experiments, while section 9 contains the conclusion.

%\begin{enumerate}
%\item \cite{chigavazira2019}: What: extending  equitable  property  division  divorce  laws  to  unmarried cohabiting couples in Australia. Findings: men increase their employment and women increase time spent on housework. Couples have more children and are more likely  to  become  home  owners. Strategy: triple Diff-in-Diff: state-time-relationship(coh/mar). They select a sample of marriages and cohabitation that stays intact at the end of the time span[this seems wrong to me]. Then they claim that the effect is not due to selection, even tought they check that in a sloppy way without a duration model, which seems the thing to do.
%\item \cite{chiappori2017}: what:  granting alimony rights to cohabiting couples in Canada. Finsings: Among couples united before the reform, obtaining the right to petition for alimony led women to lower their labor force participation but not among newly formed cohabiting couples. Cool difference lead by selection+commitment
%\item \cite{persson2020}: VERY IMPORTANT PAPER. Exploiting Sweden’ s elimin ation of s ur v ivors insur ance, I demon strate that severin g this link (1) affected entry into marriage up to 50 years before expected payout, (2) raised the divorce rate by 10pp, and (3) raised the assortativeness of matching (4) increased rate of cohabitation and of separation, consistent with selection along match quality
%
%\end{enumerate}
\section{US Divorce and Cohabitation Laws: an Overview}
\textbf{Divorce Laws.} Between late 1960s and early 1980s most U.S. states experienced fundamental changes in the divorce law, both regarding the right to initiate a divorce without the consent of the other spouse and about the laws that govern the division of assets upon divorce.

Before the 1960s the vast majority of US states had a mutual consent divorce regime,\footnote{All the states apart from New Mexico, Oklahoma and Alaska.} where the agreement of both spouses was needed to obtain a divorce for mundane reasons (i.e. without misconduct by any spouse). Moreover, divorce was permitted for grounds showing guilt of misconduct by any of the two spouses: for those cases the agreement of the innocent party alone was enough for having a divorce granted.\footnote{Examples of guilt or misconduct are adultery or abandonment.}
From the late 1960s and early 1980s most U.S. states switched to a unilateral divorce regime.\footnote{A list of the years at which unilateral divorce was introduced in different states can be found in table 1 of \cite{ciacci2017}.} Under this regime divorce can be filed by one spouse without the consent of the other.

Another dimension along which divorce laws differs across states and over time is the one about property division regimes. In the United States there are three types of those:

>>>>

\begin{abstract}
This paper analyzes the role of unilateral divorce for the rise of unmarried cohabitation. Exploiting the staggered introduction of unilateral divorce across the US states, we show that after the reform singles become more likely to cohabit than to marry, and newly formed cohabitations last longer. We then provide a theoretical rationale for these facts, building a life-cycle model with partnership choice, female labor force participation, and saving decisions. A structural estimation of the model suggests that unilateral divorce decreases couples’ stability, making cohabitation preferred to couples that would have been at high risk of divorce if they marry. As cohabiting couples formed after the reform are better matched, the average length of cohabitations increases. A counterfactual experiment reveals that the time spent cohabiting would have been halved if the divorce laws had never changed.
\end{abstract}

@@

Unmarried cohabitation is on the rise: the share of women that ever cohabited in the United States moved from 33\% in 1987 to 60\% in 2010 \citep{manning2013}. This increase contributed to the overall changes in the structure and behavior of the American family. The higher instability of cohabitation plausibly contributes to the rise in the number of single mothers (\cite{bumpass2000}), which is associated with poor outcomes for children \citep{chetty2018,mclanahan2013}. Cohabiting people spend less money on relationship-specific investments % maybe any examples here?
\citep{poortman2012}. Finally, their children's well-being is worse even after controlling for parental resources \citep{brown2004}. However, it is not clear to what extent cohabitants' outcomes are due to selection % elaborate what kind of?
versus the direct effect of the form of partnership on the couple's behavior. Quantifying these two mechanisms' relative importance can be done only by understanding the rationale for the partnerships choices. Why do people cohabit instead of marrying? Why did cohabitation become more common over time?

Our paper addresses these questions by focusing on a major US policy change that took place mostly during the 1970s. During this period, most of the states have made divorce considerably easier by switching from the mutual consent divorce regime, requiring both spouses to agree to divorce, towards unilateral divorce, in which one spouse's decision was enough to initiate the procedure. The paper explores what role does unilateral divorce have for the rise of cohabitation. Since marriage and cohabitation can be viewed as contracts whose attractiveness depends on their termination rights and costs, the switch from mutual consent to unilateral divorce represents a unique opportunity to learn about partnership choices.

We answer the questions with four contributions. First, we offer a few pieces of empirical evidence. We show that under the unilateral divorce people prefer cohabitation to marriage more often. We also argue that the newly formed cohabitations last longer. Second, we propose a theory of partnership choice and endogenous breakup/divorce. In particular, in our model the policy makes cohabitation the preferred option to couples that would have experienced the highest risk of a costly divorce. Therefore, the average match quality of cohabiting couples increases, which causes cohabitation spells to last longer. % A theory professor asked me once if something I claim to be a result is generic or is it specific to my choice of parameters. This may be a useless thing, but maybe the wording can slightly be adjusted top.
Third, we structurally estimate our model to match the empirical findings about the divorce regimes' transition. Fourth, we perform several counterfactual experiments to understand the overall role of unilateral divorce on the rise of cohabitation. We then quantify the welfare gains and losses of the reform for men and women. Thus, this paper consists of four parts, one for each contribution, which we now describe in more detail.

@@

In the first part of the paper we document the effect of unilateral divorce on the choice between marriage and cohabitation and the duration of newly formed cohabitations. We use data from the National Survey of Family and the Household (NSFH) and from the National Survey of Family Growth (NSFG) to study the choice between marrying and cohabiting. Then, exploiting the exogenous variation coming from the staggered introduction of unilateral divorce over time across the US states, we estimate that newly formed couples are 5-7\% more likely to choose cohabitation over marriage than in the pre-reform period. % Give some baseline numbers here?
Interestingly, the effect's size depends on how property is divided upon divorce, being strongest in states where each spouse gets half of the wealth and where the judge decides the assets' shares. This suggests that divorce allocations affect partnership choices.
%This suggests that the result cannot be entirely driven by divorce being a less binding commitment technology,\footnote{This mechanism is highlighted by \cite{lafortune2020}, where the equal division of assets within marriage makes it possible to efficiently invest in children even after the rise of unilateral divorce.} since the egalitarian property division rule can act as a substitute for the mutual consent regime.
Moreover, we analyze how unilateral divorce affected cohabitation spells duration: our estimates from a multinomial probit show that cohabitations formed after the reform last longer because people both marry less and break up less.\footnote{Hereafter we refer to the separation from cohabitation as a breakup to avoid confusion with legal separation.}

In the second part of the paper, we propose a theory to understand the mechanisms underlying the facts documented in the empirical part. We build a dynamic model of intra-household decision making and search in the mating market, where agents make decisions according to the realization of idiosyncratic permanent income shocks, their amount of wealth and couple-specific match quality. With some probability, single agents meet a potential partner with a random match quality, productivity, and wealth. After the draw, they decide whether to marry, cohabit, or stay single. Couples make decisions about consumption, savings, and female labor force participation. Women receive a productivity penalty for not working, and women's time can be used to produce a public good that captures utility gains from children, durable goods, and services.

In the model, cohabitation and marriage differ in their splitting costs and the way property is divided when the couple dissolves. Moreover, there is a stigma affecting cohabitations, which is modeled as an exogenous disutility flow. It can capture the shame towards out-out-wedlock births, premarital sex, and premarital cohabitation at the time of the divorce revolution. In the case of a breakup, assets are split according to individual property rights. In the case of divorce, we assume they are divided in half. We estimate the model using community property states data to be consistent with this assumption. Breakup (unilateral divorce) can be initiated unilaterally, as opposed to mutual consent divorce, which requires both partners' agreement. Following \cite{voena2015}, under mutual consent the couple always cooperates while married, which implies that the allocation of resources corresponds to the Pareto-efficient inter-temporal allocation. % Wait is this even true???
Instead, the fact that just one spouse can decide to terminate the relationship results in a lack of commitment, making the intra-household allocation of decision power responsive to shocks. Hence, abandoning the mutual consent regime affects the risk of divorce and, in turn, the surplus of marriage.

The divorce costs and rights affect the gains of marriage relative to cohabitation through three main channels. First, by acting as commitment technologies, they enforce a better risk sharing and a more efficient specialization in public goods production within marriage. We assume that public goods are produced with money and women's time. When a woman stops working to produce such a good, her potential productivity in the labor market decreases. Second, divorce costs increase the risk of being ``trapped" in a bad marriage. Third, they affect the expected value of marriage by modifying the risk and value of divorce. The effects of tightening or relaxing the barriers to divorce depend on which channel prevails. For example, the introduction of unilateral divorce has an uncertain impact on the share of couples that cohabit and marry. The outcomes of cohabitation and marriage depend not only on the rules underlying these contracts but also on sorting on unobserved relationship quality. A cheap breakup is most attractive to couples whose match quality is low. The couples' ability to cooperate depends on the match quality itself through its effect on the couples' stability: making partnership-specific investments is more comfortable when the risk of splitting is low, which is when the match quality is high.

%In our framework marriage gains depend on the quality of the match: for couples whose value is low cohabitation is the best choice because breaking up is cheaper and their risk of splitting is high. Interestingly, the ability of the couple to cooperate is directly affected by the quality of the match. For this reason, the outcomes of cohabitation and marriage depend not only of the rules underlying these contracts, but the ability of the couple to cooperate is directly affected by the quality of the match.

@@

In the third part of the paper we structurally estimate the model to understand the quantitative relevance of the mechanisms that drive partnership choice. The model is estimated by indirect inference using to match the regression results from our empirical analysis, mating market moments (NSFH), and female labor supply moments (PSID). The introduction of unilateral divorce is modeled as an unexpected policy change. The estimated model closely replicates the targeted moments. Our over-identification checks support the estimation results: the model can match several non-targeted moments, for instance, the impact of unilateral divorce on cohabitation duration.

According to the estimates, a switch from mutual consent to unilateral divorce causes couples to start cohabiting more by reducing the married couple's ability to cooperate and by increasing the likelihood of a costly divorce.\footnote{An increased likelihood of divorce can reduce the ability of the couple to cooperate by itself. Yet it also directly affects the marriage surplus by reducing the possibility of losing assets upon divorce. For example, if there was no wage uncertainty and women always participated in the labor market, unilateral divorce would affect marriage gains via the direct effect only.} Since cohabiting couples that would have married under the older regime are better matched than the average cohabiting couple, the reform increases the stability and the duration of newly formed cohabitations. To further deepen our understanding of the mechanisms, we analyze the intra-household bargaining changes that followed the policy change. We find that the average Pareto weight of cohabiting women increases because men, fearing to lose most of their assets because of divorce, convince women to cohabit instead of marrying in exchange for more power in the couple. This mechanism is specific to the divorce regime where assets are split evenly. Instead, if spouses always keep owning their assets separately, men would not need to choose cohabitation to insure their property. Consistently with empirical evidence, the impact of unilateral divorce on cohabitation likelihood is lower in the model under separate ownership. % This paragraph is still long and can be split somehow

The fourth and last part of the paper conducts a series of counterfactual experiments to understand the quantitative importance of the forces that contributed to the rise of cohabitation. To assess the role of unilateral divorce, we perform a counterfactual experiment where the unilateral divorce was never introduced. We find that people on average would have spent 1.35 years cohabiting instead of 2.26, while only 26.7\% of people would have ever cohabited instead of 34.8\%. In the second series of counterfactuals, we find that a 20\% increase in women wages and a 20\% drop in market prices of goods increase respectively by 1\% and 2\% the share of people that ever cohabited at age 39. Both effects are driven by a reduced scope for household specialization, which is better exploited within marriage. We also study various channels of how unilateral divorce affects welfare. The possibility to cohabit limits the welfare losses for men who can secure their assets, while women suffer more because of couple-specific investments like children deteriorate their value of divorcing more than for men.

\textbf{Literature.} This paper adds to three strands of the literature. First, by documenting how divorce laws influence the choice between marriage and cohabitation, we add to the existing literature that studies the effects of unilateral divorce. This policy change has been shown to affect the rate of divorce (\citealp{friedberg1998}, \citealp{wolfers2006}), female labor supply (\citealp{stevenson2008}, \citealp{voena2015}), savings (\citealp{voena2015}), marriage rates (\citealp{rasul2003,rasul2006}), children's well-being (\citealp{gruber2004}), family violence (\citealp{stevenson2006}), marriage-specific capital (\citealp{stevenson2007}), assortative mating (\citealp{reynoso2019}) and prostitution (\citealp{ciacci2017}), among the others. We complement the findings of \cite{rasul2003,rasul2006} by showing that the decrease in marriage rates after the introduction of unilateral divorce is not only driven by more people staying single, but also by more people that choose to cohabit. This suggest that marriage and cohabitation are substitutes.\footnote{Cohabitation can also be a substitute for being single or dating, as \cite{rindfuss1990} point out. Moreover, \cite{blasutto2020} and \cite{brien2006} claim that cohabitation can also be a complement for marriage, which allows the couple to learn about its match quality before making the binding decision of getting married.} Our paper builds on \cite{voena2015},  which studies how the interaction of unilateral divorce with property rights upon divorce affected married couples' household behavior. We extend her work both by considering cohabitation as an alternative relationship and by analyzing selection into partnership. This paper also extends the work of \cite{fernandez2017} by showing that not considering cohabitation as an alternative to marriage biases upwards the negative impact of unilateral divorce on men's welfare. The intuition is that men can limit the losses stemming from the increased risk of divorce by cohabiting.

Second, our paper adds to the literature that studies the choice between marriage and cohabitation. A first subset thereof has focused on identifying the gains of marriage and cohabitation, highlighting the role of commitment \citep{matouschek2008}, specialization within the couple \citep{gemici2014}, learning about match quality \citep{brien2006}, income dynamics \citep{blasutto2020}, assets \citep{lafortune2017, lafortune2020} and investment in children \citep{lundberg2015}. We extend these works by showing how an increase in the easiness of divorce decreases the couple's ability to cooperate and makes divorce allocations more relevant for partnership choices, since the likelihood of divorce increases. Consequently, the relative power of potential partners and the rules about the division of assets upon divorce become crucial. These results highlight a new role of partners' relative power and assets for partnership choices, which is analyzed within a framework that extends the theory of \cite{blasutto2020} and \cite{gemici2014} by including saving decisions.\footnote{\cite{lafortune2020} also highlight the role of assets: our model features their intuition that assets can act as a commitment technology, but our framework also allows assets to influence partnership choices via a direct effect of divorce's risk. Thanks to this mechanism, we can explain why unilateral divorce caused cohabitation to increase more in community property states than in title-based ones.}

Another subset of these papers studies the effect of changes in cohabitants' rights on partnership choices and cohabiting couples' behavior, highlighting the role of alimony rights  (\citealp{chiappori2017,gousse2018}), taxation \cite{leturcq2012} and division of assets at breakup (\citealp{fisher2012,gousse2018,chigavazira2019}). We extend this literature by showing that the introduction of unilateral divorce impacts both the choice to cohabit and cohabitation's stability, even though cohabitant's rights are not directly affected.\footnote{\cite{matouschek2008} study the effect of unilateral divorce on marriage rates using a framework where the choice between marriage and cohabitation is considered. Their framework does not predict changes to the propensity to cohabit.} Further, the effects on the intention to cohabit depends on property division rights, which indicates that partnership choices depend on divorce allocations. This evidence suggests that changes in family law should be designed considering that marriage and cohabitation are substitutes. 
%\cite{chiappori2017},\cite{persson2020}, \cite{leturcq2012}, \cite{gousse2018}
%Second, our paper expands the literature on cohabitation, showing how the risk of divorce is essential for understanding partnership choices. Different papers in the literature highlighted various gains of marriage with respect cohabitation: commitment \citep{matouschek2008}, labor specialization within the couple \citep{gemici2014}, learning about match quality \citep{brien2006}, income dynamics \citep{blasutto2020} and investment in children \citep{lafortune2020}. A paper closely related to ours is \cite{lafortune2020}, who highlight the role of assets as a collateral that enforces commitment and hence allows for optimal investment in children. While we think that their mechanism is important, it cannot be the whole story since we observe that the switch towards cohabitation is the strongest in states where assets are divided more equally, where their model would predict that the effect is the weakest. In this paper, we quantify how much of the shift is due to the erosion of marriage's collateral versus the role of the increased risk of divorce. We extend the work of \cite{gemici2014} and \cite{blasutto2020}, who develop collective dynamic models with limited commitment, introducing asset accumulation decisions, which are shown to play an important role for commitment and hence for partnership choice, especially when interacted with property rights upon divorce. Other papers highlighted how

Finally, this paper is tied to the extensive literature that studies the changes in the American household character over the last decades. Various studies explored the role of health improvements, wage distribution and dynamics, norms and technology on the rise in female labor force participation \citep{fernandez2004,greenwood2005,albanesi2016,greenwood2016}, the changes in household formation and dissolution \citep{greenwood2016,ciscato2019}, the rise in positive assortative mating \citep{fernandez2005,greenwood2016,ciscato2019} and the increase in the age at marriage \citep{santos2016}. We extend this literature by showing that the introduction of unilateral divorce implied a rise in cohabitation. Advances in the home production technology and the reduction in the gender wage gap also contributed to the rise.

Our modelling of the decision making in the couple builds on existing literature on limited commitment (\cite{kk1996},  \cite{ligon2002}, \cite{marcet2019} and \cite{pavoni2018}), which has been applied to dynamic collective models in the household by \cite{mazzocco2007}, \cite{mazzocco2013}, \cite{bayot2015}, \cite{rigas2015}, \cite{voena2015}, \cite{abraham2018}, \cite{lise2018}, \cite{low2018}, \cite{foerster2019} and \cite{reynoso2019} among others. The fact that couples cannot commit ex-ante to any possible division of marital surplus contributes to the creation of an imperfectly transferable utility environment, under which the Becker-Coase theorem does not hold (see \cite{galichon2019} and \cite{weber2018}, who applies the theorem to related environments). The imperfect transferability feature is important: according to \cite{becker1977}, divorce laws should not affect separation decisions \textit{``if all compensations between spouses were feasible and costless"}. \cite{chiappori2015} and \cite{fella2004} discuss the assumptions of the Becker-Coase theorem more.

The paper is organized as follows. Section 2 offers an overview of US divorce laws. Section 3 documents the effect of introducing unilateral divorce on partnership choices. Section 4 presents and develops the theoretical model. Section 5 describes the model's estimation, while section 6 discusses the main mechanisms of the model. Section 7 reports the results of the welfare analysis. Section 8 performs a series of counterfactual experiments, while Section 9 contains the conclusion.

%\begin{enumerate}
%\item \cite{chigavazira2019}: What: extending  equitable  property  division  divorce  laws  to  unmarried cohabiting couples in Australia. Findings: men increase their employment and women increase time spent on housework. Couples have more children and are more likely  to  become  home  owners. Strategy: triple Diff-in-Diff: state-time-relationship(coh/mar). They select a sample of marriages and cohabitation that stays intact at the end of the time span[this seems wrong to me]. Then they claim that the effect is not due to selection, even tought they check that in a sloppy way without a duration model, which seems the thing to do.
%\item \cite{chiappori2017}: what:  granting alimony rights to cohabiting couples in Canada. Finsings: Among couples united before the reform, obtaining the right to petition for alimony led women to lower their labor force participation but not among newly formed cohabiting couples. Cool difference lead by selection+commitment
%\item \cite{persson2020}: VERY IMPORTANT PAPER. Exploiting Sweden’ s elimin ation of s ur v ivors insur ance, I demon strate that severin g this link (1) affected entry into marriage up to 50 years before expected payout, (2) raised the divorce rate by 10pp, and (3) raised the assortativeness of matching (4) increased rate of cohabitation and of separation, consistent with selection along match quality
%
%\end{enumerate}
\section{US Divorce and Cohabitation Laws: an Overview} % This should be a subsection, otherwise it is a bit short...
\textbf{Divorce Laws.} Between the late 1960s and early 1980s, most US states experienced fundamental changes in the divorce law. These changes affected both the right to initiate a divorce without the other spouse's consent and about the division of assets upon divorce.
